---
title: "WCARO output 1 use case 1"
author: "Akvo - data science"
date: "11/9/2021"
output: 
  html_document:
    number_sections: TRUE
    css: akvostyle.css
    code_folding: "hide"
---

<div style="line-height: 2em;">

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)

# Library
library(here)
library(tidyverse)
library(data.table)
library(readxl)
library(sf)
library(sp)
library(rgdal)
library(maps)
library(kableExtra)
library(ggplot2)

### CIRCLES!
wp_2016_processed <- read.csv2(here("data/processed","wp_2016_processed.csv"))

# Selection of wp 2016 data
wp_2016_sub <- wp_2016_processed %>% 
  dplyr::select("ADM2", #"ADM3", "ADM4", "ADM5",
    "Water.point.Name", 
    "Latitude", "Longitude",
    "Functionality",
    "sdg_improved_source",
    "sphere_nr_people") %>% 
  unique() %>%
  mutate(sdg_improved_source = recode(
         sdg_improved_source,
         "Improved" = "improved",
         "Unimproved" = "unimproved",
         .default="unimproved")) %>%
   filter(ADM2 == "Kambia")
  
```

# Introduction 

There seem to be large discrepancies between the number of people that are reported having access to water, when different methodologies for estimating water access are used. In this overview three ways of estimating water access in the region Kambia will be shown to demonstrate this difference. The region was chosen because in a country comparison it showed large differences when choosing a different methodology. 

# Methodology
Country governments are required to report their progress on all the three key indicators of SDG and grade themselves in the JMP ladder. It is relatively easier to measure coverage by recording the number of water points that are installed, however one requirement of JMP to reach the level "basic" is that the water can be fetched in 30 minutes. This means a family should be able to make a round trip, including queuing, within 30 minutes. The general measurement of water coverage does not take population density taken into account. Countries face challenges in recording and updating their data either due to lack of resources or capacity to collect information about SDG. With this example we want to show the differences in coverage when accounting for population density. 

## Example 1: Coverage based on sphere guidelines

The following table shows the water points in the district Kambia at time of the 2016 water point mapping. The coverage is determined using the Sphere guidelines, which are guidelines to determine the number of people a specific type of water point can be expected serve, if it functions well. There are 4 categories in this guideline, mainly:  "tap", "hand pump", "open hand well" and "mechanized well". 

```{r example_1, message=FALSE, warning=FALSE, include=FALSE}
table1 <- wp_2016_sub %>%
  group_by(ADM2) %>%
  summarise(
    "Number of water points" = n(),
    "Possible coverage (nr of people)" = sum(sphere_nr_people, na.rm=TRUE)
  ) %>% 
  kbl("html") %>%
  kable_styling("striped", position = "center", font_size = 11, full_width = FALSE)

table2 <- wp_2016_sub %>%
  group_by(ADM2, sdg_improved_source) %>%
  summarise(
    "Number of water points" = n(),
    "Possible coverage (nr of people)" = sum(sphere_nr_people, na.rm=TRUE)
  ) %>% 
  kbl("html") %>%
  kable_styling("striped", position = "center", font_size = 11, full_width = FALSE)

table3 <- wp_2016_sub %>%
  group_by(ADM2, Functionality) %>%
  summarise(
    "Number of water points" = n(),
    "Possible coverage (nr of people)" = sum(sphere_nr_people, na.rm=TRUE)
  ) %>% 
  kbl("html") %>%
  kable_styling("striped", position = "center", font_size = 11, full_width = FALSE)

table4 <- wp_2016_sub %>%
   mutate(Functionality = recode(Functionality,
                                       "No - Broken down" = "Not functional",
                                       "No - Still under construction" = "Not functional",
                                       "No - Under rehabilitation" = "Not functional",
                                       "Yes – Functional (but not in use)" = "Functional",
                                       "Yes - But damaged" = "Functional",
                                       "Yes – Functional (and in use)" = "Functional")) %>%
  group_by(Functionality, sdg_improved_source) %>%
  summarise(
    "Number of water points" = n()) %>% 
  spread(sdg_improved_source, `Number of water points`) %>%
  kbl("html") %>%
  kable_styling("striped", position = "center", font_size = 11, full_width = FALSE)

# Population projections
pop_projections_city <- read.table(here("data/raw","pop_projections.csv"), sep=",", header=TRUE)
names(pop_projections_city) <- c("ADM2", "2016", "2018", "2020", "2022", "2024", "2026", "2028", "2030")

pop_projections <- pop_projections_city %>%
  mutate(across(c("2016":"2030"), ~ gsub(",", "", .))) %>%
  mutate(across(c("2016":"2030"), ~ as.numeric(.))) %>%
  mutate(City = ifelse(ADM2 %like% "City", "yes", "no")) %>%
  mutate(ADM2 = gsub(" District| City| Municipal", "", ADM2)) %>%
  group_by(ADM2) %>%
  summarise(across(c("2016":"2030"), ~ sum(.)))

# Population + improved water points
wp_coverage_ministry_pop_data_improved <- wp_2016_sub %>%
  group_by(ADM2, sdg_improved_source) %>%
  summarise(
    "Number of water points" = n(),
    "Possible coverage (nr of people)" = sum(sphere_nr_people, na.rm=TRUE)
  ) %>% 
  left_join(pop_projections %>% select(ADM2, `2016`) %>% mutate(`2016` = as.numeric(`2016`))) %>%
  mutate("Percentage of the population served (%)" = round(`Possible coverage (nr of people)`/`2016`,2)*100) %>%
  select(ADM2, sdg_improved_source, "Number of water points","Percentage of the population served (%)" ) 


wp_coverage_ministry_pop_data_functional <- wp_2016_sub %>%
   mutate(Functionality = recode(Functionality,
                                       "No - Broken down" = "Not functional",
                                       "No - Still under construction" = "Not functional",
                                       "No - Under rehabilitation" = "Not functional",
                                       "Yes – Functional (but not in use)" = "Functional",
                                       "Yes - But damaged" = "Functional",
                                       "Yes – Functional (and in use)" = "Functional")) %>%
  group_by(ADM2, sdg_improved_source, Functionality) %>%
  summarise("Number of water points" = n(),
    "Possible coverage (nr of people)" = sum(sphere_nr_people, na.rm=TRUE)) %>%
  arrange(sdg_improved_source) %>%
  left_join(pop_projections %>% select(ADM2, `2016`) %>% mutate(`2016` = as.numeric(`2016`))) %>%
  mutate("Percentage of the population served (%)" = round(`Possible coverage (nr of people)`/`2016`,2)*100) %>%
  select(ADM2, sdg_improved_source, Functionality, "Percentage of the population served (%)" ) 

wp_coverage_ministry_pop_data_functional <- wp_coverage_ministry_pop_data_functional %>%
  mutate(csum = cumsum(`Percentage of the population served (%)`))

pie_data_1 <- data.frame(methodology = c("Ministry data all water points",
                                         "Ministry data improved water points",
                                         "Ministry data improved and functional water points"),
                         percentage_served = c(82, 76, 43)) %>%
  mutate(percentage_not_served = 100 - percentage_served) %>% melt()

pie_data_1 <- pie_data_1  %>%
  mutate(methodology = str_wrap(methodology, width = 20)) %>%
  mutate(methodology = factor(methodology,
                              levels = c("Ministry data all\nwater points",
                                         "Ministry data\nimproved water\npoints",
                                         "Ministry data\nimproved and\nfunctional water\npoints")))

table5 <- wp_coverage_ministry_pop_data_functional %>%
  kbl("html") %>%
  kable_styling("striped", position = "center", font_size = 11, full_width = FALSE)


```

<br> `r table1` <br> 

As you can see from table 1 Kambia has 750 water points that have the ability to cover about 0.29 million people. Of these water points 686 are "improved", which  that have the ability to cover 0.27 million people, see table 2. 

<br> `r table2` <br> 

Table 3 shows that 750 water point 433 were functional at time of the data collection, and in table 4 you can see out of the 433, 399 were classified as "improved". 

<br> `r table3` <br> 

<br> `r table4` <br>

Based on a national inventory in 2016, there were `r pop_projections[pop_projections$ADM2 == "Kambia","2016"]` people living in Kambia in 2016, and based on projections this would have grown to `r pop_projections[pop_projections$ADM2 == "Kambia","2018"]` by 2018. If we were simply to combine the population with the possible coverage, in Kambia  `r round(291950/pop_projections[pop_projections$ADM2 == "Kambia","2016"],2)*100`% of the population is served by the water from the water points. When combining that with the type of water point 77% of the population in Kambia is served by improved water points, see table 5.

<br> `r table5` <br> 

The following graph summarizes the results from above, showing the importance of including information about the water source type and the functionality status when determining water point access.

```{r pie1, fig.align="center", fig.cap="Figure 1: Estimated basic water coverage with district level data", echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = pie_data_1, aes(x = "", y = value, fill=variable )) + 
  geom_bar(stat = "identity", position = position_fill()) +
  scale_fill_manual(values = c( "#02818a", "#a6bddb"), labels=c("Population served", "Population not served")) +
  geom_text(aes(label = paste0(value, "%")), position = position_fill(vjust = 0.5)) +
  coord_polar(theta = "y") +
  facet_wrap(.~methodology, ncol=4)  +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank()) + 
  theme(legend.position='right') + 
  guides(fill=guide_legend(nrow=2, byrow=TRUE))

```

```{r example_2, message=FALSE, warning=FALSE, include=FALSE}
# Shape file data
sl.shape <-read_sf(dsn = here::here("data/raw/SIL_admin_SHP/", "SIL.shp"))
sl.shape.data <- ggplot2::fortify(sl.shape, region='NAME') %>%
  filter(ADM2 == "Kambia")

fb_pop_sl <- read.csv2(here("data/raw", "population_sle_2019-07-01.csv"), sep=",")
fb_pop_sl <- fb_pop_sl %>% 
  mutate(ID = paste0("V", rownames(fb_pop_sl)))

fb_pop_kambia <- fb_pop_sl %>%
  mutate(Lon = as.numeric(Lon)) %>%
  mutate(Lat = as.numeric(Lat)) %>%
  mutate(Population = as.numeric(Population)) %>%
  filter(Lon > -14 & Lon < -12) %>%
  filter(Lat > 8 & Lat < 10)

FB_density_admin <- st_read(here("data/processed", "FB density with admin level.shp"))
```

This shows that only looking at the possible coverage does not tell the complete story of water supply. Adding current population numbers is critical for identifying vulnerable areas. 

Kambia is about 3000 squared kilometers and is estimated to have about 800 villages. When you take into account the requirement of the JMP that water should be collectible within a 30 minute round trip, it is important to know if the water points are dispersed is a way that fits the population density. 

The map below shows the location of the different water points in Kambia with the functionality status. The highest concentration of water points seems to be in the middle, and more to the south, and the lowest concentration is in the outer points of the district. Along the eastern boarder there seem to be quite some not functional water points. 


```{r wp_density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(sl.shape.data) +
  geom_sf(size = 0.1, color = "#603f83ff", fill="#f2f2f2") +
  theme_minimal() +
  geom_point(data=wp_2016_sub,
             aes(x=Longitude, y=Latitude, color=Functionality),
             size = 2, alpha=1, shape=20) +
  scale_color_manual(values=c("#810f7c","#a964a2","#ceaac9","#a9c4ab","#62986a","#006d2c")) +
  # theme(panel.grid.major = element_line("white")) +
  labs(title="2016 Waterpoint mapping SL: Kambia", x="", y="") +
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size=8),
        plot.title = element_text(size=14),
        axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8),
        axis.ticks = element_blank())
```

The map below shows the population density of Kambia based on the 2019 Facebook density data that determines the number of people that live on a 30 meter square are. The closer together the dots are the more dense the population, and the darker the color, the more people live within a square of 30 m2. In line with the water point density, the population seems the least concentrated in the outer points of the district, and the most concentrated in the center, and to the south. One big difference is the area on the eastern boarder in the center of the district. There seem to be living a lot of people in this area, however this is not reflected in the water point density map.

```{r pop_density, echo=FALSE, message=FALSE, warning=FALSE}
FB_density_admin_kambia <- FB_density_admin %>% 
  filter(ADM2 == "Kambia") %>%
  mutate(Population = as.numeric(Population))

ggplot() +
  geom_sf(data = FB_density_admin_kambia, aes(color=Population)) +
  theme_minimal() +
  scale_color_distiller(palette = 'RdYlBu', breaks=c(10,15,20,25,30,35)) +
  # scale_color_gradient(low="#e5f5f9", high="#2ca25f", breaks=c(10,15,20,25,30,35)) +
  labs(title="Population density in Kambia", color="Nr of people per 30m2")

```
If we want to see how many people live close enough to the water points for the water supply to be considered "basic" water service level, we need to combine the population density data with the water point locations. By taking an area around the water point and determining how many people live within that area, we can first determine the population that is close enough to the water point. Next, we will compare the amount of people to the amount of people the water point can serve based on the sphere guidelines. By looking at the intersection and comparing that to the total population we can determine the percentage of the population actually served. Below we will follow these steps one by one.  

1. Draw a circle around the water points that represents the 30 min round trip. We will use a 500m radius, based on ..

```{r circles, echo=TRUE, message=FALSE, warning=FALSE}
# Convert the water point geolocation to geometry, set the crs to EPSG:4326 (lat/long), and transform to EPSG:3035
wp_2016_sf <- st_as_sf(
  wp_2016_sub, 
  coords = c( "Longitude", "Latitude"), 
  crs = 4326) %>% 
  st_transform(3035)

# Create a buffer of a 500m radius
wp_2016_circles <- st_buffer(wp_2016_sf, dist = 500)

# Make sure the shape file has the same crs 
sl.shape <- sl.shape %>% st_set_crs(4326) %>% 
  st_transform(3035)

# Intersect water point circles with the shape file to account for overlapping water points
wp_2016_int_circles <- st_intersection(sl.shape, wp_2016_circles)

# Make sure the population data has the same crs 
pnts_sf <- st_as_sf(fb_pop_sl, 
                    coords = c( "Lon", "Lat"), 
                    crs = 4326) %>% 
  st_transform(3035)

# Identify the population that falls within the water point radius
kept_points_with_ID <- st_intersection(wp_2016_int_circles, pnts_sf) 

# Drop the geometry variable (for analysis speed) and remove possible duplicate entries
kept_points_with_ID_no_geometry <- kept_points_with_ID %>% 
  st_drop_geometry() %>%
  select(Water.point.Name, ADM2, sdg_improved_source, sphere_nr_people, ID.1, Population) %>% 
  mutate(Population = as.numeric(Population)) %>%
  unique() 

# Identify population squares that fall within reach of multiple water points -> make sure they are not counted multiple times
duplicate_pop_density <- kept_points_with_ID_no_geometry %>% 
  group_by(ID.1) %>% 
  summarise(count=n())

# divide population density number by the amount it has been counted to spread the density equally
kept_points_with_ID_no_geometry <- kept_points_with_ID_no_geometry %>% 
  left_join(duplicate_pop_density) %>%
  mutate(Population = Population/count)  

# summarise the population density per water point 
population_dens_per_wp <- kept_points_with_ID_no_geometry %>% 
  group_by(Water.point.Name, ADM2, sphere_nr_people) %>% 
  summarise(Population_within_circle = sum(Population, na.rm=TRUE)) 

population_dens_per_wp <- population_dens_per_wp %>%
  mutate(sphere_nr_people = replace_na(sphere_nr_people, 0)) %>%
  mutate(Percentage_served_within_circle = round(sphere_nr_people/Population_within_circle,2)*100) %>%
  mutate(percentage_actually_served = ifelse(Percentage_served_within_circle > 100, 100, Percentage_served_within_circle)) 

wp_2016_int_circles %>% 
  left_join(population_dens_per_wp) %>%
  ggplot() +
  geom_sf(aes(color=percentage_actually_served)) +
  theme_minimal() +
  scale_color_distiller(palette = 'YlOrRd', breaks=c(100,75,50,25,0)) +
  # scale_color_gradient(low="#e5f5f9", high="#2ca25f", breaks=c(10,15,20,25,30,35)) +
  labs(title="Percentage of the population served", color="percentage_actually_served")


```

```{r percentage_radius}

# Show percentage of population served within the 500m radius 
# and show percentage of the population that lives outside of the 500m radius
wp_coverage_fb_density <- FB_density_admin_kambia %>% 
  st_drop_geometry() %>%
  mutate(Population = as.numeric(Population)) %>%
  group_by(ADM2) %>% 
  summarise(Total_pop_fb = sum(Population)) %>%
  left_join(population_dens_per_wp %>% 
              group_by(ADM2) %>%
              summarise(sphere_nr_people = sum(sphere_nr_people, na.rm=TRUE),
                        Population_within_circle = sum(Population_within_circle, na.rm=TRUE))) %>% 
  mutate(Population_outside_circle = Total_pop_fb - Population_within_circle) %>% 
  mutate("Percentage served within 500m radius" = round(sphere_nr_people/Population_within_circle,2)*100) %>% 
  mutate(Percentage_pop_outside_circle = round(Population_outside_circle/Total_pop_fb,2)*100) %>% 
  mutate("Actual percentage served within 500m radius" = ifelse(
    `Percentage served within 500m radius`>100, 100, `Percentage served within 500m radius`)) %>%
  mutate(actual_pop_served = Population_within_circle*(`Actual percentage served within 500m radius`/100)) %>% 
  mutate("Actual percentage of the population served" = round(
   actual_pop_served/Total_pop_fb,2)*100) 

wp_coverage_fb_density %>%
  select(ADM2, Total_pop_fb, "Percentage served within 500m radius", 
         "Actual percentage served within 500m radius",
         "Actual percentage of the population served" ) %>% melt() %>% kbl("html") %>%
  kable_styling("striped", position = "left", font_size = 11)
```


## Improved status

Because of our interested in the JMP "basic" service water level, we have to take the water point status into account as well. Only "improved" water points are considered "basic". The table below shows what happens to the population statistics when we only look at improved water sources in Kambia.

```{r wp_improved, echo=FALSE, message=FALSE, warning=FALSE}

## ONLY IMPROVED
improved_kept_points_with_ID_no_geometry <- kept_points_with_ID %>% 
  st_drop_geometry() %>%
  filter(sdg_improved_source == "improved")

improved_kept_points_with_ID_no_geometry <- improved_kept_points_with_ID_no_geometry %>% 
  
  select(Water.point.Name, ADM2, sphere_nr_people, ID.1, Population) %>% 
  mutate(Population = as.numeric(Population)) %>%
  unique() 

# Duplicates:
improved_duplicate_pop_density <- improved_kept_points_with_ID_no_geometry %>% 
  group_by(ID.1) %>% 
  summarise(count=n())

improved_kept_points_with_ID_no_geometry <- improved_kept_points_with_ID_no_geometry %>% 
  left_join(improved_duplicate_pop_density) %>%
  mutate(Population = Population/count)  

improved_kept_points_with_ID_no_geometry <- improved_kept_points_with_ID_no_geometry %>% 
  group_by(Water.point.Name, ADM2, sphere_nr_people) %>% 
  summarise(Population_within_circle = sum(Population, na.rm=TRUE)) 

# Show percentage of population served within the 500m radius 
# and show percentage of the population that lives outside of the 500m radius
improved_wp_coverage_fb_density <- FB_density_admin_kambia %>% 
  st_drop_geometry() %>%
  mutate(Population = as.numeric(Population)) %>%
  group_by(ADM2) %>% 
  summarise(Total_pop_fb = sum(Population)) %>%
  left_join(improved_kept_points_with_ID_no_geometry %>% 
              group_by(ADM2) %>%
              summarise(sphere_nr_people = sum(sphere_nr_people, na.rm=TRUE),
                        Population_within_circle = sum(Population_within_circle, na.rm=TRUE))) %>% 
  mutate(Population_outside_circle = Total_pop_fb - Population_within_circle) %>% 
  mutate("Percentage served within 500m radius" = round(sphere_nr_people/Population_within_circle,2)*100) %>% 
  mutate(Percentage_pop_outside_circle = round(Population_outside_circle/Total_pop_fb,2)*100) %>% 
  mutate("Actual percentage served within 500m radius" = ifelse(
    `Percentage served within 500m radius`>100, 100, `Percentage served within 500m radius`)) %>%
  mutate(actual_pop_served = Population_within_circle*(`Actual percentage served within 500m radius`/100)) %>% 
  mutate("Actual percentage of the population served" = round(
   actual_pop_served/Total_pop_fb,2)*100) 


improved_wp_coverage_fb_density %>%
  select(ADM2, Total_pop_fb, "Percentage served within 500m radius", 
         "Actual percentage served within 500m radius",
         "Actual percentage of the population served" ) %>% melt() %>% kbl("html") %>%
  kable_styling("striped", position = "left", font_size = 11)

```

## Functionality status 

While households under the category basic are relatively better off than the earlier ones, it is important to check the availability of water from the listed sources. The data here needs to be disaggregated by ‘functionality’ i.e. ask whether the source is working and providing the adequate quantity of water as per the national/ international standards. The table below shows what happens to the population statistics when we only look at improved and functional water sources in Kambia.

```{r wp_functional, echo=FALSE, message=FALSE, warning=FALSE}

## ONLY IMPROVED & FUNCTIONAL

functional_kept_points_with_ID_no_geometry <- kept_points_with_ID %>% 
  st_drop_geometry() %>%
  filter(sdg_improved_source == "improved") %>%
  filter(Functionality %in% c("Yes – Functional (and in use)", 
                              "Yes – Functional (but not in use)",
                              "Yes - But damaged" )) %>% 
  select(Water.point.Name, ADM2, sdg_improved_source, sphere_nr_people, ID.1, Population) %>% 
  mutate(Population = as.numeric(Population)) %>%
  unique() 

# Duplicates:
functional_duplicate_pop_density <- functional_kept_points_with_ID_no_geometry %>% 
  group_by(ID.1) %>% 
  summarise(count=n())

functional_kept_points_with_ID_no_geometry <- functional_kept_points_with_ID_no_geometry %>% 
  left_join(functional_duplicate_pop_density) %>%
  mutate(Population = Population/count)  

functional_kept_points_with_ID_no_geometry <- functional_kept_points_with_ID_no_geometry %>% 
  group_by(Water.point.Name, ADM2, sphere_nr_people) %>% 
  summarise(Population_within_circle = sum(Population, na.rm=TRUE)) 

# Show percentage of population served within the 500m radius 
# and show percentage of the population that lives outside of the 500m radius
functional_wp_coverage_fb_density <- FB_density_admin_kambia %>% 
  st_drop_geometry() %>%
  mutate(Population = as.numeric(Population)) %>%
  group_by(ADM2) %>% 
  summarise(Total_pop_fb = sum(Population)) %>%
  left_join(functional_kept_points_with_ID_no_geometry %>% 
              group_by(ADM2) %>%
              summarise(sphere_nr_people = sum(sphere_nr_people, na.rm=TRUE),
                        Population_within_circle = sum(Population_within_circle, na.rm=TRUE))) %>% 
  mutate(Population_outside_circle = Total_pop_fb - Population_within_circle) %>% 
  mutate("Percentage served within 500m radius" = round(sphere_nr_people/Population_within_circle,2)*100) %>% 
  mutate(Percentage_pop_outside_circle = round(Population_outside_circle/Total_pop_fb,2)*100) %>% 
  mutate("Actual percentage served within 500m radius" = ifelse(
    `Percentage served within 500m radius`>100, 100, `Percentage served within 500m radius`)) %>%
  mutate(actual_pop_served = Population_within_circle*(`Actual percentage served within 500m radius`/100)) %>% 
  mutate("Actual percentage of the population served" = round(
   actual_pop_served/Total_pop_fb,2)*100) 

functional_wp_coverage_fb_density %>%
  select(ADM2, Total_pop_fb, "Percentage served within 500m radius", 
         "Actual percentage served within 500m radius",
         "Actual percentage of the population served" ) %>% melt() %>% kbl("html") %>%
  kable_styling("striped", position = "left", font_size = 11)
```


## Functionality status PLUS

The table below shows what happens to the statistics when we only consider improved, functional water points that are not damaged and in use at time of the data collection:


```{r wp_functional_plus, echo=FALSE, message=FALSE, warning=FALSE}

## ONLY IMPROVED & FUNCTIONAL

functional_plus_kept_points_with_ID_no_geometry <- kept_points_with_ID %>% 
  st_drop_geometry() %>%
  filter(sdg_improved_source == "improved") %>%
  filter(Functionality %in% c("Yes – Functional (and in use)")) %>% 
  select(Water.point.Name, ADM2, sdg_improved_source, sphere_nr_people, ID.1, Population) %>% 
  mutate(Population = as.numeric(Population)) %>%
  unique() 

# Duplicates:
functional_plus_duplicate_pop_density <- functional_plus_kept_points_with_ID_no_geometry %>% 
  group_by(ID.1) %>% 
  summarise(count=n())

functional_plus_kept_points_with_ID_no_geometry <- functional_plus_kept_points_with_ID_no_geometry %>% 
  left_join(functional_plus_duplicate_pop_density) %>%
  mutate(Population = Population/count)  

functional_plus_kept_points_with_ID_no_geometry <- functional_plus_kept_points_with_ID_no_geometry %>% 
  group_by(Water.point.Name, ADM2, sphere_nr_people) %>% 
  summarise(Population_within_circle = sum(Population, na.rm=TRUE)) 

# Show percentage of population served within the 500m radius 
# and show percentage of the population that lives outside of the 500m radius
functional_plus_wp_coverage_fb_density <- FB_density_admin_kambia %>% 
  st_drop_geometry() %>%
  mutate(Population = as.numeric(Population)) %>%
  group_by(ADM2) %>% 
  summarise(Total_pop_fb = sum(Population)) %>%
  left_join(functional_plus_kept_points_with_ID_no_geometry %>% 
              group_by(ADM2) %>%
              summarise(sphere_nr_people = sum(sphere_nr_people, na.rm=TRUE),
                        Population_within_circle = sum(Population_within_circle, na.rm=TRUE))) %>% 
  mutate(Population_outside_circle = Total_pop_fb - Population_within_circle) %>% 
  mutate("Percentage served within 500m radius" = round(sphere_nr_people/Population_within_circle,2)*100) %>% 
  mutate(Percentage_pop_outside_circle = round(Population_outside_circle/Total_pop_fb,2)*100) %>% 
  mutate("Actual percentage served within 500m radius" = ifelse(
    `Percentage served within 500m radius`>100, 100, `Percentage served within 500m radius`)) %>%
  mutate(actual_pop_served = Population_within_circle*(`Actual percentage served within 500m radius`/100)) %>% 
  mutate("Actual percentage of the population served" = round(
   actual_pop_served/Total_pop_fb,2)*100) 

functional_plus_wp_coverage_fb_density %>%
  select(ADM2, Total_pop_fb, "Percentage served within 500m radius", 
         "Actual percentage served within 500m radius",
         "Actual percentage of the population served" ) %>% melt() %>% kbl("html") %>%
  kable_styling("striped", position = "left", font_size = 11)
```


```{r visuals1, echo=FALSE, message=FALSE, warning=FALSE}

pie_data <- wp_coverage_ministry_pop_data_improved %>% 
  filter(sdg_improved_source == "improved") %>%
  select(ADM2, `Percentage of the population served (%)`) %>%
  rename(percentage_served = `Percentage of the population served (%)`) %>% 
  mutate(percentage_not_served = 100 - percentage_served) %>% melt() %>%
  mutate(methodology = "District level population data improved sources") %>% 
  full_join(
    wp_coverage_fb_density %>% 
      select(ADM2, "Actual percentage of the population served") %>%
      rename(percentage_served = `Actual percentage of the population served`) %>% 
      mutate(percentage_not_served = 100 - percentage_served) %>% melt() %>%
      mutate(methodology = "Population density data (30m2)")) %>% 
  full_join(
    improved_wp_coverage_fb_density %>% 
      select(ADM2, "Actual percentage of the population served") %>%
      rename(percentage_served = `Actual percentage of the population served`) %>% 
      mutate(percentage_not_served = 100 - percentage_served) %>% melt() %>%
      mutate(methodology = "Population density data (30m2) improved sources")) %>% 
  full_join(
    functional_wp_coverage_fb_density %>% select(ADM2, "Actual percentage of the population served") %>%
      rename(percentage_served = `Actual percentage of the population served`) %>% 
      mutate(percentage_not_served = 100 - percentage_served) %>% melt() %>%
      mutate(methodology = "Population density data (30m2) improved and functional sources"))

pie_data <- pie_data  %>%
  mutate(methodology = str_wrap(methodology, width = 20)) %>%
  mutate(methodology = factor(methodology, 
                              levels = c("District level\npopulation data\nimproved sources",
                                         "Population density\ndata (30m2)",
                                         "Population density\ndata (30m2) improved\nsources",
                                         "Population density\ndata (30m2) improved\nand functional\nsources")))

ggplot(data = pie_data, aes(x = "", y = value, fill=variable )) + 
  geom_bar(stat = "identity", position = position_fill()) +
  scale_fill_manual(values = c( "#02818a", "#a6bddb"), labels=c("Population served", "Population not served")) +
  geom_text(aes(label = paste0(value, "%")), position = position_fill(vjust = 0.5)) +
  coord_polar(theta = "y") +
  facet_wrap(.~methodology, ncol=4)  +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank()) + 
  theme(legend.position='right') + 
  guides(fill=guide_legend(nrow=2, byrow=TRUE))

```


</div>